---
title: "APSTA-GE: 2003 Intermediate Quantitative Methods"
subtitle: "Lab Section 003, Week 12"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  pdf_document:
    highlight: kate
    latex_engine: xelatex
mainfont: Helvetica
---

```{r setup, include=FALSE}
# Created on: 11/24/2020
# Modified on: 11/29/2020
# Author: Tong Jin
# Affiliation: New York University
# Contact: tj1061@nyu.edu

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

# Dependencies
library(stringr)
library(MASS)
```

# Introduction {.tabset}

In this lab, we will examine various factors that may be associated with student's math performance. The attached dataset, `student-mat.csv`, is a random sample of student's math scores in secondary education of two Portuguese schools. It was collected by using school reports and quesitonnaires. 

Note: the dataset is in `csv` format with semi-colon as separators. 

```{r load-data}
dat <- read.csv("student-mat.csv", sep = ";", header = TRUE)
```

## Dimensions

```{r dim}
dim(dat)
table(is.na(dat))
```

The dataset contains 395 records with 33 columns. No data point is missing.

### Variables

Here is a list of column names.

```{r var}
as.data.frame(colnames(dat))
```

### Distributions

Plot distribution of numeric variables.

#### Math Score (Label)

Range: 0 - 20

```{r var-G3-summary}
summary(dat$G3)
```

```{r var-G3}
hist(dat$G3)
```

## Regression Models

### Baseline models

```{r mod_BL}
mod_BL <- lm(G3 ~ 1, data = dat)
summary(mod_BL)
```

### Y ~ Age

```{r mod_age}
mod_age <- lm(G3 ~ age, data = dat)
summary(mod_age)
```

### Y ~ age + sex

```{r mod_age_sex}
mod_age_sex <- lm(G3 ~ age + factor(sex), data = dat)
summary(mod_age_sex)
```

### Model with All Predictors

```{r mod_full}
for (col in 1:ncol(dat)) {
  if (class(dat[, col]) == "character") {
    dat[, col] <- as.factor(dat[, col])
  }
}
mod_full <- lm(G3 ~ ., data = dat)
summary(mod_full)
```

## Model Selection: Step-wise AIC

### Additive Terms Only

Start from the full model. Run stepAIC by going "backward". You can also use the baseline model and go "forward" or use any model and go with "both" directions.

```{r step-AIC}
stepAIC(mod_full, scope = . ~ 1, direction = "backward")
```

This is the selected model. It has the lowest AIC score (best prediction performance).

```{r}
mod_lowest_AIC <- lm(G3 ~ school + age + activities + romantic + famrel + Walc + 
                     absences + G1 + G2, data = dat)
summary(mod_lowest_AIC)
```

### with Interactive Terms

```{r}
# stepAIC(mod_full, scope = . ~ .^2, direction = "both")
```

---
END
---