---
title: "APSTA-GE 2003: Intermediate Quantitative Methods"
subtitle: "Lab Section 003, Week 10"
institute: "New York University"
date: "11/10/2020"
output:
  xaringan::moon_reader:
    css: ["styles.css", "styles-font.css"]
    self_contained: TRUE
    seal: TRUE
    lib_dir: libs
    nature:
      ratio: '16:10'
      highlightStyle: github
      highlightLines: TRUE
      countIncrementalSlides: FALSE
---

```{r setup, include=FALSE}
# Author: Tong Jin
# Date created: 11/03/2020
# Date modified: 11/03/2020

knitr::opts_chunk$set(
  echo = TRUE, 
  message=FALSE, 
  warning=FALSE, 
  fig.height = 6,
  fig.width = 10
)

# Dependencies
library(plotly)
library(dplyr)
library(DT)
library(grid)
library(gridExtra)
library(stringr)
library(corrplot)
```

## Reminders
  
- Sample Assignment Solutions Available
  - In [this Google Drive folder](https://drive.google.com/drive/folders/1b3fHkxC81ov2Sw8uPomXpXHyar2w2eQu?usp=sharing)

- No Assignment Due
  - <(￣︶￣)>
- Office hours
  - Monday 9:00 - 10:00am (EST)
  - Wednesday 12:30 - 1:30pm (EST)
  - Additional time slots available
    - Sign-up sheet [HERE](https://docs.google.com/spreadsheets/d/1YY38yj8uCNIm1E7jaI9TJC494Pye2-Blq9eSK_eh6tI/edit?usp=sharing)
  
- Office hour Zoom link [HERE](https://nyu.zoom.us/j/9985119253)
  
- Office hour notes
  - Available on NYU Classes

---

## Today's Topics

- Analyzing NYC Property Sales Data using Multiple Regression Modeling

---

## Introduction

In this lab, we will explore a property market dataset using multiple regression modeling. We will examine how the sale price varied across a 12-month period in 2016 - 2017. Specifically, we will try to understand the covariates that affect the sale price. Our research questions are:

1. How did the sale price of NYC properties vary in 2016 - 2017? 

2. What contributed to the increase of the sale price?

3. How do we predict the sale price based on the columns we have?

## Design

We will conduct an Exploratory Data Analysis (EDA), followed by a multiple regression analysis, to try to answer these questions.

---

## Dataset

The dataset records every building or building unit (apartment, etc.) sold in the New York City property market over a 12-month period in 2016 - 2017. (Source: [Kaggle](https://www.kaggle.com/new-york-city/nyc-property-sales))

The dataset contains the location, address, type, sale price, and sale data of building units sold. 

[Dataset file here](https://drive.google.com/file/d/1JjV_UZW6eCeAuqyTY3yV5W3WZdZ9ve2A/view?usp=sharing)

First, load the data into R: 

```{r load-data}
dat <- read.csv("../nyc-rolling-sales.csv", header = TRUE, 
                na.strings = c("NA", " -  ")) #<<
dim(dat)
```

Number of observations: 84,548

Number of columns: 22

Column Names: 

```{r dis-column-names}
colnames(dat)
```

---

## Codebook - Part 1

(Source: [Glossary of Terms for Property Sales Files - NYC](https://www1.nyc.gov/assets/finance/downloads/pdf/07pdf/glossary_rsf071607.pdf))

- X: integer: index
- BOROUGH: integer: borough name
  - 1: Manhattan
  - 2: Bronx
  - 3: Brooklyn
  - 4: Queens
  - 5: Staten Island
- NEIGHBORHOOD: character: neighborhood name
- BUILDING.CLASS.CATEGORY: category: identify similar properties by broad usage (e.g. One Family Homes) without looking up individual Building Classes.
- TAX.CLASS.AT.PRESENT: category: tax classes
  - Class 1: residential property of up to three units
  - Class 2: all other properties that are primarily residential
  - Class 3: property owned by a gas, telephone or electric company
  - Class 4: all other commercial properties
- BLOCK: integer: sub-division of the borough
- LOT: integer: a tax lot
- EASE.MENT: logical: easement or not
- BUILDING.CLASS.AT.PRESENT: category: constructive use
- ADDRESS: character: address
- APARTMENT.NUMBER: character: apt. number
- ZIP.CODE: category: zip code
- RESIDENTIAL.UNITS: integer: number of residential units
- COMMERCIAL.UNITS: integer: number of commercial units
- TOTAL.UNITs: integer: total number of units
- LAND.SQUARE.FEET: character: land sqft
- GROSS.SQUARE.FEET: character: gross sqft
- YEAR.BUILT: integer: year built
- BUILDING.CLASS.AT.TIME.OF.SALE: category: constructive use
- SALE.PRICE: character: sale price
- SALE.DATE: character: sale date

---

## Pre-process

First, make column names to lowercase using the `stringr` package. Then, replace `.` with `_`.

```{r pre}
colnames(dat) <- str_to_lower(colnames(dat)) #<<
colnames(dat) <- str_replace_all(colnames(dat), pattern = "[.]", replacement = "_") #<<
colnames(dat)
```

---

## Pre-process

- Examine if there is any missing value.

```{r describe}
n_col <- ncol(dat)
paste("Column Index", "Number of NAs", sep = ", ")
for (index in 1:n_col) {
  print(paste(index, ", ", colnames(dat)[index], ", ", sum(is.na(dat[, index])), sep = ""))
}
```

---

## Trim the dataset

Drop unrelated columns. 

```{r trim}
col_name_to_drop <- c("x", "ease_ment", "neighborhood", 
                      "building_class_at_present", "tax_class_at_present", 
                      "address", "apartment_number", "tax_class_at_time_of_sale")
col_index_to_drop <- which(colnames(dat) %in% col_name_to_drop)
dat <- dat[, -col_index_to_drop]
dim(dat)
```

Drop rows with missing values.

```{r trim1}
col_name_w_NA <- c("sale_price", "land_square_feet", "gross_square_feet")
col_index_w_NA <- which(colnames(dat) %in% col_name_w_NA)
row_w_NA <- which(is.na(dat[, col_index_w_NA[1]]))
dat <- dat[-row_w_NA, ]
row_w_NA <- which(is.na(dat[, col_index_w_NA[2]]))
dat <- dat[-row_w_NA, ]
row_w_NA <- which(is.na(dat[, col_index_w_NA[3]]))
dat <- dat[-row_w_NA, ]
dim(dat)
table(is.na(dat))
```

--- 

## Descriptive Analysis

```{r descriptive}
colnames(dat)
```

```{r sale-date}
dat$sale_date <- as.Date(dat$sale_date)
```


---

## Borough

```{r borough}
p <- ggplot(dat, aes(x = factor(borough))) + 
  geom_histogram(stat="count") + 
  scale_x_discrete(labels = c("Man", "Bronx", "Brooklyn", "Queens", "SI")) + 
  theme_minimal()
ggplotly(p)
```

---

## Average Sale Price Per Borough

```{r borough-price}
average_sales_per_borough <- dat %>% 
  group_by(borough) %>%
  summarize(total = n(), average_sales = mean(sale_price))
average_sales_per_borough
```

---

## Viz

```{r borough-price-viz}
p <- ggplot(average_sales_per_borough, aes(x = borough, y = average_sales)) +
  geom_point() + 
  scale_x_discrete(labels = c("Man", "Bronx", "Brooklyn", "Queens", "SI")) + 
  theme_minimal()
ggplotly(p)
```

---

## Building Class Category

```{r class}
building_class_cat <- dat %>%
  group_by(borough, building_class_category) %>%
  summarize(total = n()) %>% 
  arrange(desc(total), borough)
building_class_cat
```

---

## Viz

```{r class-viz}
p <- ggplot(building_class_cat, aes(x = building_class_category, y = total)) +
  geom_point() + 
  theme_classic()
ggplotly(p)
```

---

## Zip code

```{r zip}
zip_code <- dat %>%
  group_by(zip_code) %>%
  summarize(average_sales = mean(sale_price)) %>%
  arrange(desc(average_sales))
zip_code
```

---

## Viz 

```{r zip-viz}
p <- ggplot(zip_code, aes(x = factor(zip_code), y = average_sales)) +
  geom_point() + 
  theme_classic()
ggplotly(p)
```

---

## Other Columns

```{r land-sqft}
summary(dat$land_square_feet)
summary(dat$year_built)
summary(dat$total_units)
```

---

## Check correlation

```{r actual-corr}
cor_dat <- cor(dat[, -c(2, 12, 14)])
corrplot::corrplot(cor_dat, type = "lower") #<<
```

---

## Let's Build Our Model Together!

## Contact

Tong Jin

- Email: tj1061@nyu.edu
- Office Hours
  - Mondays, 9 - 10am (EST)
  - Wednesdays, 12:30 - 1:30pm (EST)

