---
title: "APSTA-GE 2003: Intermediate Quantitative Methods"
subtitle: "Lab Section 003, Week 12"
institute: "New York University"
date: "11/24/2020"
output:
  xaringan::moon_reader:
    css: ["styles.css", "styles-font.css"]
    self_contained: TRUE
    seal: TRUE
    lib_dir: libs
    nature:
      ratio: '16:10'
      highlightStyle: github
      highlightLines: TRUE
      countIncrementalSlides: FALSE
---

```{r setup, include=FALSE}
# Author: Tong Jin
# Date created: 11/24/2020
# Date modified: 11/24/2020

knitr::opts_chunk$set(
  echo = TRUE, 
  message=FALSE, 
  warning=FALSE, 
  fig.height = 6,
  fig.width = 10
)

# Dependencies
```

## Reminders

- Assignment 7 ([group project](https://drive.google.com/drive/folders/1S0XC8s6-N4OM4X158DmAWjpxtCl_sVRg?usp=sharing))
  - [Group assignment](https://docs.google.com/spreadsheets/d/1uZ_ao7dts_wiipddpM9wz_JSYWfBFvcvZ5r6wvqkjTs/edit?usp=sharing)
  - Due: 12/08/2020 (Friday) 5:00 PM
  - Sample solutions will be available after due

- Office hours
  - Monday 9:00 - 10:00am (EST)
  - Wednesday 12:30 - 1:30pm (EST)
  - Additional time slots available
    - Sign-up sheet [HERE](https://docs.google.com/spreadsheets/d/1YY38yj8uCNIm1E7jaI9TJC494Pye2-Blq9eSK_eh6tI/edit?usp=sharing)
  - Office hour Zoom link [HERE](https://nyu.zoom.us/j/9985119253)

---

## Today's Topics

- Predict student' math scores
- Model Selection with Step-wise AIC

---

## The dataset

**Student Performance Data Set**

**Source**: 

[Machine Learning Repository, UCI](https://archive.ics.uci.edu/ml/datasets/Student+Performance#)

Paulo Cortez, University of Minho, Guimarães, Portugal, http://www3.dsi.uminho.pt/pcortez

```{r load-data}
dat <- read.csv("../student-mat.csv", sep = ";", header = TRUE)
dim(dat)
table(is.na(dat))
```

---

## Variables

```{r colnames}
colnames(dat)
```

Label: `G3`

Features: Everything else

---

## Plan

1. EDA

2. Fit regression models

3. Evaluate model bias and variance

---

## Prediction

The goal of prediction using regression is to fit a model that has the lowest AIC score.

---

## Bias-variance tradeoff

[Bias-variance tradeoff](https://en.wikipedia.org/wiki/Bias%E2%80%93variance_tradeoff)

---

## Contact

Tong Jin

- Email: tj1061@nyu.edu
- Office Hours
  - Mondays, 9 - 10am (EST)
  - Wednesdays, 12:30 - 1:30pm (EST)

