---
title: "APSTA-GE 2003: Intermediate Quantitative Methods"
subtitle: "Lab Section 003, Week 11"
institute: "New York University"
date: "11/17/2020"
output:
  xaringan::moon_reader:
    css: ["styles.css", "styles-font.css"]
    self_contained: TRUE
    seal: TRUE
    lib_dir: libs
    nature:
      ratio: '16:10'
      highlightStyle: github
      highlightLines: TRUE
      countIncrementalSlides: FALSE
---

```{r setup, include=FALSE}
# Author: Tong Jin
# Date created: 11/03/2020
# Date modified: 11/03/2020

knitr::opts_chunk$set(
  echo = TRUE, 
  message=FALSE, 
  warning=FALSE, 
  fig.height = 6,
  fig.width = 10
)

# Dependencies
library(plotly)
library(dplyr)
library(DT)
library(grid)
library(gridExtra)
library(stringr)
library(corrplot)
library(MASS)
library(lmtest)
```

## Reminders

- Assignment 6
  - Due: 11/20/2020 (Friday) 11:55 PM
  - Sample solutions will be available after due
    - In [this Google Drive folder](https://drive.google.com/drive/folders/1b3fHkxC81ov2Sw8uPomXpXHyar2w2eQu?usp=sharing)

- Office hours
  - Monday 9:00 - 10:00am (EST)
  - Wednesday 12:30 - 1:30pm (EST)
  - Additional time slots available
    - Sign-up sheet [HERE](https://docs.google.com/spreadsheets/d/1YY38yj8uCNIm1E7jaI9TJC494Pye2-Blq9eSK_eh6tI/edit?usp=sharing)
  - Office hour Zoom link [HERE](https://nyu.zoom.us/j/9985119253)

---

## Today's Topics

- Walk through Assignment 6
- Model Selection: NYC Property Sales Data

---
class: inverse, center, middle

# Walk through Assignment 6

---

## Question 4

*Fit a regression assuming a quadratic relationship between income and age by further including a squared age term in the regression model.*

Quadratic Model: 

$$\text{Sales} = \beta_0 + \beta_1 \cdot \text{YouTube} + \beta_2 \cdot \text{YouTube}^2 + \varepsilon$$

where $\beta_0$ is the intercept, $\beta_1$ is the coefficient of YouTube marketing budget, $\beta_2$ is the coefficient of squared YouTube marketing budget, and $\varepsilon$ is the residual.

---

## Question 6

*At what two consecutive ages does the expected change in income (from one age to the next) change signs (i.e, from positive to negative or vice versa) as compared to the prior two consecutive ages?*

- The stationary point of a **concave** curve is the point with maximum y value.
- The stationary point of a **convex** curve is the point with minimum y value.

Options: 

- eyeball: predict and plot
- manual calculation: use vertex form
- get every possible y value: use a loop

---

## Question 10

*Based on the regression of log(income)~age above, we expect average income...*

Logarithm ~ proportional

---

## Part 2 Question 1

*Run a regression model predicting the y variable with x1 (gender) and x2 (group) as predictors (use their appropriate dummy/factor representations).*

A regression model with a quantitative dependent variable and two categorical independent variables.

**For example:** D.V.: test score as `score`. 

I.V.: study team as `team`, biological sex as `sex`.

- `team` has three groups: `A`, `B`, `C`
  - `team A`: 0
  - `team B`: 1
  - `team C`: 2
- `sex` has two groups: `male`, `female`
  - `female`: 0
  - `male`: 1

**Understanding the reference group:**

In the model, we set the coefficient of the reference group as **zero**. In other words, we treat reference groups as constants. Because they are constants, we add them into the intecept.

---

## P2Q1 Demo

```{r dgp-P2Q1}
set.seed(111)
dgp <- function (n, beta0, beta1, beta2) {
  team_groups <- c(0, 1, 2)
  error <- rnorm(n, mean = 0, sd = 10)
  team <- factor(sample(team_groups, size = n, replace = TRUE), levels = c(0, 1, 2))
  sex <- factor(sample(c(0, 1), size = n, replace = TRUE), levels = c(0, 1))
  score <- beta0 + beta1 * as.numeric(team) + beta2 * as.numeric(sex) + error
  dat <- data.frame(score, team, sex)
  summary(lm(score ~ team + sex, data = dat))
}
(df <- dgp(n = 100, beta0 = 40, beta1 = 2, beta2 = 1.1))
```

$$\text{score} = 45.0466 - 2.439 \cdot \text{team}_1 + 0.6065 \cdot  \text{team}_2 + 2.6817 \cdot \text{sex}_{\text{male}}$$

---

## Part 2 Question 4 (Q16)

Q: How to change reference groups? 

A: `relevel()`

---
class: inverse, center, middle

# Model Selection: NYC Property Sales Data

---

## Introduction

**Purpose:** examine the covariates that affect the sale price.

**Research questions**:

1. How did the sale price of NYC properties vary in 2016 - 2017? 

2. What contributed to the increase of the sale price?

3. How do we predict the sale price based on the columns we have?

**Data**: records every building or building unit (apartment, etc.) sold in the New York City property market over a 12-month period in 2016 - 2017. (Source: [Kaggle](https://www.kaggle.com/new-york-city/nyc-property-sales))

```{r load-data}
dat <- read.csv("../nyc-rolling-sales.csv", header = TRUE, na.strings = c("NA", " -  "))
dim(dat)
```

Number of observations: 84,548

Number of columns: 22

---

## Codebook

(Source: [Glossary of Terms for Property Sales Files - NYC](https://www1.nyc.gov/assets/finance/downloads/pdf/07pdf/glossary_rsf071607.pdf))

- X: integer: index
- BOROUGH: integer: borough name
- NEIGHBORHOOD: character: neighborhood name
- BUILDING.CLASS.CATEGORY: category: identify similar properties by broad usage (e.g. One Family Homes) without looking up individual Building Classes.
- TAX.CLASS.AT.PRESENT: category: tax classes
  - Class 1: residential property of up to three units
  - Class 2: all other properties that are primarily residential
  - Class 3: property owned by a gas, telephone or electric company
  - Class 4: all other commercial properties
- BLOCK: integer: sub-division of the borough
- LOT: integer: a tax lot
- EASE.MENT: logical: easement or not
- BUILDING.CLASS.AT.PRESENT: category: constructive use
- ADDRESS: character: address
- APARTMENT.NUMBER: character: apt. number
- ZIP.CODE: category: zip code
- RESIDENTIAL.UNITS: integer: number of residential units
- COMMERCIAL.UNITS: integer: number of commercial units
- TOTAL.UNITs: integer: total number of units
- LAND.SQUARE.FEET: character: land sqft
- GROSS.SQUARE.FEET: character: gross sqft
- YEAR.BUILT: integer: year built
- BUILDING.CLASS.AT.TIME.OF.SALE: category: constructive use
- SALE.PRICE: character: sale price (**Label (Dependent variable)**)
- SALE.DATE: character: sale date

---

## Pre-process (Done)

Let's assume we fixed columns names, dropped unrelated columns and rows with missing values.

```{r pre}
n_col <- ncol(dat)
# Fix column names
colnames(dat) <- str_to_lower(colnames(dat))
colnames(dat) <- str_replace_all(colnames(dat), pattern = "[.]", replacement = "_")
# Drop unrelated columns
col_name_to_drop <- c("x", "ease_ment", "neighborhood", 
                      "building_class_at_present", "tax_class_at_present", 
                      "address", "apartment_number", "tax_class_at_time_of_sale")
col_index_to_drop <- which(colnames(dat) %in% col_name_to_drop)
dat <- dat[, -col_index_to_drop]
# Drop rows with missing values
col_name_w_NA <- c("sale_price", "land_square_feet", "gross_square_feet")
col_index_w_NA <- which(colnames(dat) %in% col_name_w_NA)
row_w_NA <- which(is.na(dat[, col_index_w_NA[1]]))
dat <- dat[-row_w_NA, ]
row_w_NA <- which(is.na(dat[, col_index_w_NA[2]]))
dat <- dat[-row_w_NA, ]
row_w_NA <- which(is.na(dat[, col_index_w_NA[3]]))
dat <- dat[-row_w_NA, ]
dat$sale_date <- as.Date(dat$sale_date)
```

---

## Remove Extreme Values

It's best practice to remove outliers before fitting regression models.

```{r extreme-value}
summary(dat$sale_price)
row_index_to_drop <- which(dat$sale_price > 3000000)
dat <- dat[-row_index_to_drop, ]
```

---

## Visualize sale prices

```{r sale-price-viz}
p <- ggplot(dat, aes(y = sale_price)) + 
  geom_boxplot() + 
  theme_minimal()
ggplotly(p)
```

---

## Descriptive Analysis

Dependent variable (Label): `sale_price`

Independent variable (Features, predictors):

```{r features}
colnames(dat)[-which(colnames(dat) == "sale_price")]
```

---

## Check correlation

In real world, it is rare that independent variables are actually **independent**. There are always correlations among IVs.

```{r actual-corr}
index_categorical_columns <- c(2, 12, 14)
cor_dat <- cor(dat[, -index_categorical_columns])
corrplot::corrplot(cor_dat, type = "lower")
```

---

## Remove highly-correlated IVs and building classes

The strategy here is to remove highly correlated IVs.

```{r remove-highly-corr}
col_name_to_drop <- c("residential_units", "commercial_units", "gross_square_feet", 
                      "building_class_category", "building_class_at_time_of_sale")
col_index_to_drop <- which(colnames(dat) %in% col_name_to_drop)
dat <- dat[, -col_index_to_drop]
```

---

## Get Ready for Modeling

Label: `sale_price`

Features: 

```{r features-ready}
colnames(dat)[-which(colnames(dat) == "sale_price")]
```

---

## Baseline Model

Baseline model: **mean average**: 0.54 million

```{r baseline}
mod_BL <- lm(sale_price ~ 1, data = dat)
summary(mod_BL)
```

---

## Full Model

$$
\begin{aligned}
\text{sale price} & = 5026000 - 80140 \cdot \text{borough} - 26.2 \cdot \text{block} - 35.66 \cdot \text{lot} - 25.6 \cdot \text{zip code} \\ 
& - 381.4 \cdot \text{total units} + 764.9 \cdot \text{land sqft} + 55.35 \cdot \text{year built} + 29.31 \cdot \text{sale date}
\end{aligned}
$$

```{r full}
mod_full <- lm(sale_price ~ ., data = dat)
summary(mod_full)
```

---

## How to select features?

How to select features = regression model selection = examine **goodness of fit**

Methods: 

1. R-squared

2. Adjust R-squared

3. **Log Likelihood**

4. Akaike information criterion (**AIC**)

5. Bayesian information criterion (BIC)

---

## Likelihood Ratio Test

Likelihood: How likely the model simulate the variance of population data (sample mean is equal to the population mean).

Log likelihood: log of the likelihood. The higher log likelihood, the higher the likelihood.

$$\lambda = \left( 1 + \frac{t^2}{n - 1} \right)^{-\frac{n}{2}}$$

---

## Demo: Log Lik

```{r loglik}
lrtest(mod_full)
```

---

## AIC

Akaike information criterion

$$\text{AIC} = 2k - 2ln(L) = 2k + n \cdot ln(\frac{\text{RSS}}{n})$$

where $k$ is the number of estimated parameters and $L$ is the maximum value of the likelihood function for the model.

---

## Step-wise AIC

`stepAIC(mod_full, scope = list(lower = mod_BL, upper = mod_full), direction = "backward")`

`stepAIC(mod_BL, scope = list(lower = mod_BL, upper = mod_full), direction = "forward")`

---

## Contact

Tong Jin

- Email: tj1061@nyu.edu
- Office Hours
  - Mondays, 9 - 10am (EST)
  - Wednesdays, 12:30 - 1:30pm (EST)

