---
title: "APSTA-GE 2003: Intermediate Quantitative Methods"
subtitle: "Lab Section 003, Week 7"
institute: "New York University"
date: "10/20/2020"
output:
  xaringan::moon_reader:
    css: ["styles.css", "styles-font.css"]
    self_contained: TRUE
    seal: TRUE
    lib_dir: libs
    nature:
      ratio: '16:10'
      highlightStyle: github
      highlightLines: TRUE
      countIncrementalSlides: FALSE
---

```{r setup, include=FALSE}
# Author: Tong Jin
# Date created: 10/20/2020
# Date modified: 10/20/2020

knitr::opts_chunk$set(
  echo = TRUE, 
  message=FALSE, 
  warning=FALSE, 
  fig.height = 6,
  fig.width = 10
)

# Dependencies
library(plotly)
library(openintro)
library(DT)
library(grid)
library(gridExtra)

set.seed(1111)
```

## Reminders

- **Group Assignment**

  - Solutions available [HERE](https://drive.google.com/drive/folders/17GnapKdf_PzsHo1s0vMCx_gFZIq7v5qE?usp=sharing)
  - Comments on group submission available soon
  
- **Assignment 4**

  - Due: **10/28/2020 11:55pm (EST)**
  
- Office hours

  - Monday 9 - 10am (EST)
  - Wednesday 12:30 - 1:30pm (EST)
  - Additional time slots available
    - Sign-up sheet [HERE](https://docs.google.com/spreadsheets/d/1YY38yj8uCNIm1E7jaI9TJC494Pye2-Blq9eSK_eh6tI/edit?usp=sharing)
  
- Office hour Zoom link
  - https://nyu.zoom.us/j/97347070628 (pin: 2003)
  
- Office hour notes
  - Available on NYU Classes
- Updates on the shiny app soon

---

## Today's Topics

- Review on Regression Analysis

  - Influential vs. Leverage Points
  
- (Slightly) Review on Multiple Regression
    
- Regression Simulation

  - Using Simulator on Shiny App

---
class: inverse, center, middle

# Review Regression Analysis

---

## Influential vs. Leverage Points

- How to measure distance?
    - **Outlier**: distance between the point and the **regression line**
    - **Leverage**: 
      - the x value of the point is far from the data
      - distance between the point and the regression line is not far
    - **Influential**:
      - the x value of the point is far from the data
      - distance between the point and the regression line is far

---

## Visualize Leverage Points

- **Linear equation**: $Y_i = \beta_1 + \beta_2 \cdot X_i$

Add a leverage point.

```{r vis-leverage, echo=FALSE}
# Generate simulated data
n <- 50
X <- rnorm(n, mean = 3, sd = 1)
dat <- data.frame(X, Y = 20 + 3 * X + rnorm(n,0,1))       # Linear model

# Fit a lm model
mod_lin <- lm(Y ~ X, data = dat)

# Add outliers
lev_point <- c(10, 2.824 * 10 + 10.11)
dat_lev <- rbind(dat, lev_point)

# Visualize different data distributions
p_lev <- ggplot(dat_lev, aes(X, Y)) + 
  geom_point(col = "red") + 
  theme_minimal()
# Plot graphs
ggplotly(p_lev)
```

---

## Check Effects on Leverage Points

```{r vis-leverage-lm-line, echo=FALSE}
p_lev <- p_lev +
  geom_smooth(data = dat, method = "lm", se = FALSE, color = "blue") + 
  geom_smooth(data = dat_lev, method = "lm", se = FALSE, color = "green") + 
  labs(title = "Original vs. Leverage Lines")
ggplotly(p_lev)
```

---

## Visualize Influential Points

- **Linear equation**: $Y_i = \beta_1 + \beta_2 \cdot X_i$

Add an influential point.

```{r vis-influential, echo=FALSE}
# Add outliers
inf_point <- c(10, 2.824 * 10 - 5)
dat_inf <- rbind(dat, inf_point)

# Visualize different data distributions
p_inf <- ggplot(dat_inf, aes(X, Y)) + 
  geom_point(col = "red") + 
  theme_minimal()
# Plot graphs
ggplotly(p_inf)
```

---

## Check Effects on Influential Points

```{r vis-influential-lm-line, echo=FALSE}
p_inf <- p_inf +
  geom_smooth(data = dat, method = "lm", se = FALSE, color = "blue") + 
  geom_smooth(data = dat_inf, method = "lm", se = FALSE, color = "green") +
  labs(title = "Original vs. Influential Lines")
ggplotly(p_inf)
```

---

## Compare Visualization on Leverage and Influential

```{r compare-lev-vs-inf, echo=FALSE}
grid.arrange(p_lev, p_inf, ncol = 2)
```

---

## Detecting Influential Points

Locate outliers by their index numbers

```{r demo-detection}
library(car)
plot(mod_lin, which = 1) #<<
```

---

## How to Remove Outliers from Data?

```{r dat-full, echo=FALSE}
dat_full <- rbind(dat, lev_point, inf_point)
```


```{r rm-outliers}
DT::datatable(dat_full, options = list(pageLength = 5))
summary(dat_lev)
```

---

## How to Remove Outliers from Data? (Continued)

```{r rm-outliers-2}
# Option 1:
dat_clean   <- dat_full[-c(51, 52), ]
# Option 2:
dat_clean_2 <- dat_full[c(1:50), ]

table(dat_clean == dat_clean_2)
```

---

class: inverse, center, middle

# Review on Multiple Regression

---

## Multiple Regression

- Regression with two or more independent variables

- To interpret, control one or more independent variables

- Equation: $$Y = \beta_0 + \beta_1 \cdot X_1 + \beta_2 \cdot X_2 + ... + \beta_i \cdot X_i + \varepsilon$$

- Dummy variables:

    - Regression cannot accept categorical variables directly
    - Convert to numeric variables

---

## Demo: Simple Regression Model

```{r demo-multi-reg}
dat_demo <- datarium::marketing
# Fit a simple regression model
summary(mod_sing <- lm(sales ~ youtube, data = dat_demo)) #<<
```

---

## Demo: Multiple Regression Model

```{r demo-multi-reg-2}
# Fit a multiple regression model
summary(mod_multi <- lm(sales ~ youtube + facebook + newspaper, data = dat_demo)) #<<
```

---

## Demo: Comparison

```{r demo-compare-models}
mod_sing$coefficients
mod_multi$coefficients
```

---

class: inverse, center, middle

# Regression Simulation

---

## Regression Simulation

**Let's move to the Shiny App.**

---

## Contact

Tong Jin

- Email: tj1061@nyu.edu
- Office Hours
  - Mondays, 9 - 10am (EST)
  - Wednesdays, 12:30 - 1:30pm (EST)

