---
title: "APSTA-GE 2003: Intermediate Quantitative Methods"
subtitle: "Lab Section 003, Week 4"
author: "Tong Jin"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  ioslides_presentation:
    incremental: false
    widescreen: true
    smaller: true
    logo: _images/logo_nyu.png
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(plotly)
library(openintro)
library(data.table)
```

## Reminders

- Assignment 2
  - Due: **10/02/2020 11:55pm (EST)**
- Office hours
  - Monday 9 - 10am (EST)
  - Wednesday 12:30 - 1:30pm (EST)
- Office hour Zoom link
  - https://nyu.zoom.us/j/97347070628 (pin: 2003)
- Office hour notes
  - Available on NYU Classes under the "Resources" tab
  
## Fun slide

> - **"What do you call a regression model with only two data points?"**

> - A line.

## Let's build the fun part

**Let's create a simulated data set with only two data points.**

```{r fun-1, echo=TRUE}
first_point  <- c(25, 10)
second_point <-  c(5, 30)
dat_for_fun <- data.frame(
  X = c(first_point[1], second_point[1]),
  Y = c(first_point[2], second_point[2])
)
dat_for_fun
```

## Here comes the fun part

**Let's fit a regression model and talk about the perfect R-squared and those NAs for half an hour.**

```{r fun-2}
# Let's fit a linear regression model
lm_for_fun <- lm(formula = Y ~ X, data = dat_for_fun)
summary(lm_for_fun)
```

## More fun? Yes!

**To make things even better, let's plot the results!**

```{r fun-3}
ggplotly(
  ggplot(data = dat_for_fun, aes(x = X, y = Y)) +
    geom_point() + 
    geom_smooth(method = lm, se = FALSE) +
    theme_minimal() + xlim(0, 30) + ylim(0, 35) + 
    geom_text(aes(label = paste("Point ", rownames(dat_for_fun), sep = "")), check_overlap = TRUE, nudge_y = 2.5, size = 3)
)
```

## How about this?

**How about four data points?**

```{r fun-4}
third_point <- c(30, 20)
fourth_point <- c(10, 15)
dat2_for_fun <- rbind(dat_for_fun, third_point)
dat2_for_fun <- rbind(dat2_for_fun, fourth_point)
avg_X <- mean(dat2_for_fun$X)
avg_Y <- mean(dat2_for_fun$Y)
p_for_fun <- ggplot(data = dat2_for_fun, aes(x = X, y = Y)) +
  geom_point()
ggplotly(
  p_for_fun +
    geom_smooth(method = lm, se = FALSE) + 
    geom_vline(xintercept = avg_X, color = "red", size = .2) + 
    geom_hline(yintercept = avg_Y, color = "red", size = .2) +
    theme_minimal() + xlim(0, 30) + ylim(0, 35) + 
    geom_text(aes(label = paste("Point ", rownames(dat2_for_fun), sep = "")), check_overlap = TRUE, nudge_y = 2.5, size = 3)
)
```

## Let's check the result

*(25, 10), (10, 15), (5, 30), (30, 20)*

```{r fun-5}
lm2_for_fun <- lm(formula = Y ~ X, data = dat2_for_fun)
summary(lm2_for_fun)
```

## Seamlessly switching to today's topics

**Distance to the mean**: the distance between each data point and the average DV value.

```{r fun-6}
ggplotly(
  p_for_fun +
    geom_smooth(method = lm, se = FALSE) + 
    geom_hline(yintercept = avg_Y, color = "red", size = .2) +
    theme_minimal() + xlim(0, 30) + ylim(0, 35) + 
    geom_text(aes(label = paste("Point ", rownames(dat2_for_fun), sep = "")), check_overlap = TRUE, nudge_x = -2.5, size = 3) + 
    geom_segment(aes(x = 5, y = 30, xend = 5, yend = avg_Y, color = "green"), linetype = "dotted", size = 0.3) + 
    geom_segment(aes(x = 10, y = 15, xend = 10, yend = avg_Y, color = "green"), linetype = "dotted", size = 0.3) +
    geom_segment(aes(x = 25, y = 10, xend = 25, yend = avg_Y, color = "green"), linetype = "dotted", size = 0.3) +
    geom_segment(aes(x = 30, y = 20, xend = 30, yend = avg_Y, color = "green"), linetype = "dotted", size = 0.3) 
)
```

## Sum of Squares (SS)

What is **Sum of Squares**?

The sum of squared distance. 

$$
\sum_{i=1}^n \ \ \ (Y_i - \overline{Y}) \ \ \ ^2
$$

## Three types of SS

**Total Sum of Squares (TSS)**: Sum of squared distance between observed data and the mean

$$
TSS = \sum_{i=1}^n (Y_i - \overline{Y})^2
$$

**Model (Explained) Sum of Squares (MSS)**: Sum of squared distance included in the model

$$
MSS = \sum_{i=1}^n (\hat{Y_i} - \overline{Y_i})^2
$$

**Residual Sum of Squares (RSS)**: Sum of squared distance not included in the model

$$
RSS = \sum_{i=1}^n e_{i}^2 = \sum_{i=1}^n (Y_i - \hat{Y_i})^2
$$

## Total Sum of Squares (TSS)

$$
TSS = \sum_{i=1}^n (Y_i - \overline{Y})^2
$$

```{r TSS}
ggplotly(
  p_for_fun +
    geom_smooth(method = lm, se = FALSE) + 
    geom_hline(yintercept = avg_Y, color = "red", size = .2) +
    theme_minimal() + xlim(0, 30) + ylim(0, 35) + 
    geom_text(aes(label = paste("Point ", rownames(dat2_for_fun), sep = "")), check_overlap = TRUE, nudge_x = -2.5, size = 3) + 
    geom_segment(aes(x = 5, y = 30, xend = 5, yend = avg_Y, color = "green"), linetype = "dotted", size = 0.3) + 
    geom_segment(aes(x = 10, y = 15, xend = 10, yend = avg_Y, color = "green"), linetype = "dotted", size = 0.3) +
    geom_segment(aes(x = 25, y = 10, xend = 25, yend = avg_Y, color = "green"), linetype = "dotted", size = 0.3) +
    geom_segment(aes(x = 30, y = 20, xend = 30, yend = avg_Y, color = "green"), linetype = "dotted", size = 0.3) 
)
```

## Total Sum of Squares (TSS) cont.

$$
\begin{align}
TSS_{fun} & = \sum_{i=1}^4 (Y_i - \overline{Y})^2 \\
          & = (10 - 18.75)^2 + (15 - 18.75)^2 + (30 - 18.75)^2 + (20 - 18.75)^2 \\
          & = 218.75
\end{align}
$$

## Model Sum of Squares (MSS)

$$
MSS = \sum_{i=1}^n (\hat{Y_i} - \overline{Y_i})^2
$$

```{r MSS}
# Predict Y_hat
Y_hat <- predict(lm2_for_fun, new = dat2_for_fun)
ggplotly(
  p_for_fun +
    geom_smooth(method = lm, se = FALSE) + 
    geom_hline(yintercept = avg_Y, color = "red", size = .2) +
    theme_minimal() + xlim(0, 30) + ylim(0, 35) + 
    geom_text(aes(label = paste("Point ", rownames(dat2_for_fun), sep = "")), check_overlap = TRUE, nudge_x = -2.5, size = 3) + 
    geom_segment(aes(x = 5, y = avg_Y, xend = 5, yend = Y_hat[2], color = "green"), linetype = "dotted", size = 0.3) + 
    geom_segment(aes(x = 10, y = avg_Y, xend = 10, yend = Y_hat[4], color = "green"), linetype = "dotted", size = 0.3) +
    geom_segment(aes(x = 25, y = avg_Y, xend = 25, yend = Y_hat[1], color = "green"), linetype = "dotted", size = 0.3) +
    geom_segment(aes(x = 30, y = avg_Y, xend = 30, yend = Y_hat[3], color = "green"), linetype = "dotted", size = 0.3) 
)
```

## Model Sum of Squares (MSS) cont.

$$
\begin{align}
MSS_{fun} & = \sum_{i=1}^4 (\hat{Y_i} - \overline{Y_i})^2 \\
          & = (15.88235 - 18.75)^2 + (23.52941 - 18.75)^2 + (13.97059 - 18.75)^2 + (21.61765 - 18.75)^2 \\
          & = 62.1323529412
\end{align}
$$

## Residual Sum of Squares (RSS)

$$
RSS = \sum_{i=1}^n e_{i}^2 = \sum_{i=1}^n (Y_i - \hat{Y_i})^2
$$

```{r RSS}
ggplotly(
  p_for_fun +
    geom_smooth(method = lm, se = FALSE) + 
    geom_hline(yintercept = avg_Y, color = "red", size = .2) +
    theme_minimal() + xlim(0, 30) + ylim(0, 35) + 
    geom_text(aes(label = paste("Point ", rownames(dat2_for_fun), sep = "")), check_overlap = TRUE, nudge_x = -2.5, size = 3) + 
    geom_segment(aes(x = 5, y = 30, xend = 5, yend = Y_hat[2], color = "green"), linetype = "dotted", size = 0.3) + 
    geom_segment(aes(x = 10, y = 15, xend = 10, yend = Y_hat[4], color = "green"), linetype = "dotted", size = 0.3) +
    geom_segment(aes(x = 25, y = 10, xend = 25, yend = Y_hat[1], color = "green"), linetype = "dotted", size = 0.3) +
    geom_segment(aes(x = 30, y = 20, xend = 30, yend = Y_hat[3], color = "green"), linetype = "dotted", size = 0.3) 
)
```

## Residual Sum of Squares (RSS) cont.

$$
\begin{align}
RSS_{fun} & = \sum_{i=1}^4 (Y_i - \hat{Y_i})^2 \\
          & = (10 - 15.88235)^2 + (30 - 23.52941)^2 + (20 - 13.97059)^2 + (15 - 21.617656)^2 \\
          & = 156.617732353
\end{align}
$$

## TSS, MSS, and RSS

In our case:

$$
TSS_{fun} = 218.75 \\
MSS_{fun} = 62.1323529412 \\
RSS_{fun} = 156.617732353
$$

$$
TSS_{fun} = 218.75\\
MSS_{fun} + RSS_{fun} = 62.1323529412 + 156.617732353 = 218.750085294 \\ 
TSS_{fun} = MSS_{fun} + RSS_{fun}
$$

## Analysis of Variance 

```{r ANOVA, echo=TRUE}
anova(lm2_for_fun)
```

$$
F_{fun} = \frac{MSS_{fun}/(p-1)}{RSS_{fun}/(n-p)} = \frac{62.132}{156.618/(4 - 2)} = 0.793
$$

where $p$ is the degrees of freedom.

## R Squared

$R^2$ is the percentage of variability in the data set explained by the model.

$$
R^2 = \frac{MSS}{TSS} = 1 - \frac{RSS}{TSS}
$$

In our case:

$$
\begin{align}
R_{fun}^2 & = \frac{MSS_{fun}}{TSS_{fun}} = 1 - \frac{RSS_{fun}}{TSS_{fun}} \\
          & = \frac{62.132}{218.75} = 1 - \frac{156.618}{218.75} \\
          & = 0.284
\end{align}
$$

