---
title: "APSTA-GE 2003: Intermediate Quantitative Methods"
subtitle: "Sample Solution - Assignment 2"
output:
  pdf_document:
    latex_engine: xelatex
    highlight: tango
geometry: margin=1in
header-includes:
   - \usepackage{amsmath,amsthm}
   - \usepackage{amssymb}
mainfont: Helvetica
sansfont: Times New Roman
monofont: Hack
fontsize: 10pt
---

**Created on:** 11/09/2020

**Modified on:** `r format(Sys.time(), '%m/%d/%Y')`

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

# Author: Tong Jin
# Affiliation: New York University
# Contact: tj1061@nyu.edu
```

## Instructions

In this section, you will use the data set `parent_son.csv` to conduct regression analysis and answer questions.

Throughout this homework (all parts), please round all numerical answers to the nearest 100th decimal point, unless otherwise instructed in the question.

```{r load data}
# Load the dataset `parent_son.csv` to R using read.csv(), 
# define the dataset as `dat`.
dat <- read.csv("../data/parent_son.csv")

# Check the structure of `dat`
str(dat)
```

First, load the dataset to R using `read.csv()` as shown above. Then, check dimensions and the structure:

`dat` has 1,078 rows and 4 columns: 

- `id`: Index
- `fheight`: Father's height, numeric
- `sheight`: Son's height, numeric
- `mheight`: Mother's height, numeric

```{r preset}
# Create a function to round to the nearest 100th decimal point
rd <- function (n) {
  round(as.numeric(n), digits = 2)
}
```

\newpage

## Question 1

Estimate a simple linear regression using mother's height ("`mheight`") as independent variable X, and son's height ("`sheight`") as outcome variable Y.

Use R function: `lm(Y ~ X, data)`

The intercept is: ____

The slope is: ____

### Answer: Q1

```{r Q1}
mod_Q1 <- lm(sheight ~ mheight, data = dat)
summary(mod_Q1)

intercept_mod_Q1 <- 47.56386
slope_mod_Q1 <- 0.32566
```

The intercept is `r rd(intercept_mod_Q1)`.

The slope is `r rd(slope_mod_Q1)`.

<br>

## Question 2

In your own language, explain what the intercept means (in the context of mother and son's height).

### Answer: Q2

The intercept is the average height of sons whose mother's heights are 0 inch. 

<br>

## Question 3

In your own language, explain the meaning of the slope in the context of this analysis.

### Answer: Q3

For two groups of mothers whose heights differ by one inch, the taller group will have sons who are 0.32 inches taller on average.

<br>

## Question 4

Report the 95% confidence interval of the slope coefficient

lower bound: ____

upper bound: ____

### Answer: Q4

```{r Q4}
confint(mod_Q1)

lwr_bound <- 0.3021704
upp_bound <- 0.349154
```

lower bound: `r rd(lwr_bound)`.

upper bound: `r rd(upp_bound)`.

<br>

## Question 5

What is the expected height of sons whose mothers' heights are 65 inches tall? (keep two decimal point)

### Answer: Q5

We can manually predict by using the model equation: 

$$
\begin{aligned}
\text{Son's Height}      & = 47.56386 + 0.32566 \cdot \text{Mother's Height} \\
\text{Mother's Height}_1 & = 65                                              \\
\text{Son's Height}_1    & = 47.56386 + 0.32566 \cdot 65 = 68.73
\end{aligned}
$$

Alternatively, we can use `predict()`. 

```{r Q5}
new_dat <- data.frame(mheight = 65)
predict(mod_Q1, newdata = new_dat)
```

<br>

## Question 6

Estimate the average height of sons whose mother's height is 60 inches and the confidence interval of this estimate.

**Hint:** you can use R command `predict()`. Use "?predict.lm" to learn about how "predict" works for "lm" objects.

Average height of sons whose mother's height is 60 inches: ____

The lower bound of the 95% CI: ____

The upper bound of the 95% CI: ____

### Answer: Q6

```{r Q6}
new_dat <- data.frame(mheight = 60)
predict(mod_Q1, newdata = new_dat, interval = "confidence")
```

Average height of sons whose mother's height is 60 inches: 67.1.

The lower bound of the 95% CI: 66.93.

The upper bound of the 95% CI: 67.28.

<br>

## Question 7

Maria is 60 inches tall, can you predict her son's adult height based on the regression model you estimated? Prescribe a 95% prediction interval of your prediction.

Predicted height: ____

Lower bound of the 95% prediction interval: ____

Upper bound of the 95% prediction interval: ____

### Answer: Q7

```{r Q7}
predict(mod_Q1, newdata = new_dat, interval = "prediction")
```

Since we are referring to the same regression model, the predicted value will be the same. However, in this question, our prediction is based on the information of a single individual instead of the average, the prediction interval (individual-level) will be wider than the confidence interval (average-level) to reflect reduced accuracy.

Predicted height: 67.1.

Lower bound of the 95% prediction interval: 62.85.

Upper bound of the 95% prediction interval: 71.36.

<br>

## Question 8

Please explain briefly why the prediction interval is considerably wider in Question 7 than the confidence interval in Question 6.

You can offer intuitions or use mathematical formula for the explanation.

### Answer: Q8

In Question 6, the confidence interval reflects the uncertainty around an estimated mean height of a group, while in Question 7, the prediction interval reflects the uncertainty around a predicted value of an individual. It is harder to predict individual's height than estimate the mean height of a group.

```{r Q8-viz}
# Visualize the difference between individual and average
par(mfrow = c(1, 2)) # Create a 1-by-2 grid canvas

# Draw confidence interval
mheight <- sort(dat$mheight, decreasing = FALSE)
new_dat <- data.frame(mheight)
sheight <- predict(mod_Q1, newdata = new_dat, interval = "confidence")
plot(dat$mheight, dat$sheight, type = "n") # pch: point character
title(main = "Confidence Interval")
polygon(x = c(rev(mheight), mheight), 
        y = c(rev(sheight[, "upr"]), sheight[, "lwr"]), 
        col = "light blue", border = NA)
abline(a = 47.56386, b = 0.32566, col = "red")
abline(v = 60, lty = "dashed", col = "blue")
points(x = 60, y = 66.93, pch = 16, col = "blue", cex = .5)
points(x = 60, y = 67.28, pch = 16, col = "blue", cex = .5)

# Draw prediction interval
sheight <- predict(mod_Q1, newdata = new_dat, interval = "prediction")
plot(dat$mheight, dat$sheight, type = "n") # pch: point character
title(main = "Prediction Interval")
polygon(x = c(rev(mheight), mheight), 
        y = c(rev(sheight[, "upr"]), sheight[, "lwr"]), 
        col = "light blue", border = NA)
abline(a = 47.56386, b = 0.32566, col = "red")
abline(v = 60, lty = "dashed", col = "blue")
points(x = 60, y = 62.85, pch = 16, col = "blue", cex = .5)
points(x = 60, y = 71.36, pch = 16, col = "blue", cex = .5)
```

<br>

## Question 9

The R squared is the percentage of the variability in Y explained by X. It can be calculated using the formula above.

Report the R squared: ____

### Answer: Q9

```{r Q9}
mod_Q9 <- mod_Q1
summary(mod_Q9)$r.square
```

The R squared is 0.41.

<br>

## Question 10

Report the total sum of squares (TSS) for the dependent variable (son's height).

Total sum of square: ____

### Answer: Q10

```{r Q10}
mod_Q10 <- mod_Q1
anova(mod_Q10)
```

$$
TSS = MSS + RSS = 3476.7 + 5055.9 = 8532.6
$$

<br>

## Question 11

Report the model sum of squares (MSS).
 
Model Sum of Squares: ____

### Answer: Q11

```{r Q11}
mod_Q11 <- mod_Q1
anova(mod_Q11)
```

$$
MSS = 3476.7
$$

<br>

## Question 12

Report the residual sum of squares (RSS).

Residual Sum of Squares: ____

### Answer: Q12

```{r Q12}
mod_Q12 <- mod_Q1
anova(mod_Q12)
```

$$
RSS = 5055.9
$$

<br>

## Question 13

Report model sum of squares divided by the total sum of squares (MSS/TSS).

### Answer: Q13

$$
MSS/TSS = \frac{3476.7}{8532.6} = 0.4074608
$$

The $\frac{MSS}{TSS}$ is `r rd(0.4074608)`.

<br>

## Question 14

The model sum of squares captures the part of the variability in son’s height (DV) that is due to their mother’s height variability.

### Answer: Q14

This is true.

---- 
**END:** Sample Solution - Assignment 2
----